Generation of TypeScript Declaration Files from JavaScript Code

### 2020-08-27 programming-2021/main.tex
in section 6

Our selection of type features covered in the tool is driven by
examination of the files in
the DefinitelyTyped repository. We parsed each declaration file with
\texttt{dts-parse} and tagged it according to the features used in the file.
Table~\ref{tab:dts-parse-stats} summarizes the results.
We decided to prioritize the features with the highest number of
occurrences and support them first in our framework.

@@

For each module, we applied \texttt{dts-compare} to the generated declaration file and the
one uploaded to the DefinitelyTyped repository for the same module, as shown in
\figref{fig:evaluation-diagram}. While this approach does not provide an absolute measure
of quality, it gives us at least an indication of the pragmatics of \texttt{dts-generate}:
The files on DefinitelyTyped are perceived to be useful by the community. If the accuracy
of the generated files is comparable with the accuracy of the files on DefinitelyTyped,
then \texttt{dts-generate} is a viable alternative to manually created definition files.

@@

The first step is to parse the declaration files, which can be
achieved by using the TypeScript Compiler API, a library which traverse the
Abstract Syntax Tree of a TypeScript program
\cite{typescript-compiler-api}. 
The step also performs a sanity check of the generated declaration
files as it rejects files with syntactic or semantics errors.
The output of \texttt{dts-parse} is a structure where declared
{interfaces}, {functions}, {classes} and
{namespaces} are stored separately. Function arguments are
correctly described, identifying complex types like union types or
callbacks. Optional parameters are also identified. For
{classes}, a distinction between the constructor and methods
is made. Declaration files are tagged according to their features, which is useful for identifying and filtering out declaration files containing unimplemented features.

\texttt{dts-parse} is itself written in TypeScript. It receives the file as input
parameter and returns an AST serialized in JSON, as shown in \coderef{code:dts-parse-example}.

\begin{lstlisting}[
  caption={Example for \texttt{dts-parse} executed on module \texttt{glob-to-regexp} and combined with \texttt{jq} to browse the returned JSON},
  language=bash,
label=code:dts-parse-example,
  float=tp, captionpos=b
]
$ dts-parse -i glob-to-regexp/index.d.ts | jq '.parsing.functions[0].name'
"GlobToRegexp"
\end{lstlisting} %$

\subsection{dts-compare}
\label{sec:dts-compare}
The tool is written in TypeScript. It takes two declaration files as input and returns a
JSON file containing the result of the comparison, as shown in
\coderef{code:dts-compare-example}. Following the naming convention of any standard Unit
Testing framework, the input files are marked as \lstinline{expected} or
\lstinline{actual}. We use \lstinline{expected} for the DefinitelyTyped
declaration files.

@@

\texttt{dts-compare} applies \texttt{dts-parse} internally to both files to get 
normalized structures that are easy to compare. The result of the comparison is an array
of an abstract entity \texttt{Difference}.

We extracted code examples written by
developers to execute the code and gather run-time information. The quality of the
generated declaration file is heavily dependent on the code examples. For example, if a
property of an interface does not get accessed due to a code example not exploring a
particular execution path, that property will not exist in the generated declaration
file.

To measure the impact of incomplete code examples on the
quality of generated declaration files we categorized each difference as follows: 
\begin{itemize}
  \item \texttt{solvable}: Differences that can be resolved by adding more code
    examples. For example, invoking a function with different parameters so that a missing
    interface property gets accessed. 
  \item \texttt{unsolvable}: Differences that cannot be resolved with further examples.
    % are marked as \texttt{unsolvable}. These differences are implementation errors of \texttt{dts-generate}.
\end{itemize}

The concrete evaluated differences are as follows:
\begin{itemize}
  \item \texttt{TemplateDifference}: A difference between the choice of TypeScript
    templates. For example, if the file in DefinitelyTyped is written as a \texttt{module}
    but we generated the file using the \texttt{module-function} template. The comparison
    stops if the templates differ. 
  \item \texttt{ExportAssignmentDifference}: An equality check between the export
    assignments of both files, that is in the expression \lstinline{export = XXX}. 
  \item \texttt{FunctionMissingDifference}: A function declaration is present in the expected
    file but not in the generated one. This difference applies to functions as well as methods of classes
    and interfaces. 
  \item \texttt{FunctionExtraDifference}: A function is present in the generated declaration file but not in DefinitelyTyped.
  \item \texttt{FunctionOverloadingDifference}: The number of declarations for the same
    function is different. This difference can be due to inexperience of the author of a
    declaration file, who uses, e.g., multiple declarations where a union typed argument
    would be appropriate.
  \item \texttt{ParameterMissingDifference}: A parameter of a function or a property of an interface is not present in the generated file.
  \item \texttt{ParameterExtraDifference}: A parameter of a function or a property of an
    interface is present in the generated file but not in the DefinitelyTyped file. 
  \item \texttt{ParameterTypeDifference}: A parameter of a function or a property of an interface is generated with a different type than in the DefinitelyTyped file. Here we differentiate between \texttt{SolvableDifference} and \texttt{UnsolvableDifference}.
  \begin{itemize}
    \item \texttt{SolvableDifference}: A type difference that can be solved by writing
      further code examples. For example, a basic type that is converted to a union type,
      a function overloading, or a parameter or property that is marked as optional. 
    \item \texttt{UnsolvableDifference}: Any difference not considered as \texttt{SolvableDifference}.
  \end{itemize}
\end{itemize}

@@

@@

\paragraph*{Extraction of Code Examples}
In this step, we loose another 50\% of modules! This loss is mainly
explained because some developers do not wrap their code in a block
using the \texttt{javascript} or \texttt{js} tags. However, as we are
still left with code examples for 2200 modules, we did not further
look into code extraction as this number was considered sufficient for
evaluating the generation of declaration files.

>>>>

The selection of typing features covered in the tool is driven by
examination of the files in
the DefinitelyTyped repository. We parsed each declaration file with
\texttt{dts-parse} and tagged it according to the features used in the file.
Table~\ref{tab:dts-parse-stats} summarizes the results.
We decided to prioritize the features with the highest number of
occurrences and support them first in our framework.

@@

For each module, we applied \texttt{dts-compare} to the generated declaration file and the
one uploaded to the DefinitelyTyped repository for the same module as shown in
\figref{fig:evaluation-diagram}. While this approach does not provide an absolute measure
of quality, it gives us at least an indication of the pragmatics of \texttt{dts-generate}:
The files on DefinitelyTyped are perceived to be useful by the community. If the accuracy
of the generated files is comparable with the accuracy of the files on DefinitelyTyped,
then \texttt{dts-generate} is a viable alternative to manually created definition files.

@@

The first step is to parse the declaration files, which can be
achieved by using the TypeScript Compiler API, a library which traverses the
Abstract Syntax Tree of a TypeScript program
\cite{typescript-compiler-api}. 
The step also performs a sanity check of the generated declaration
files as it rejects files with syntactic or semantics errors.
The output of \texttt{dts-parse} is a structure where declared
{interfaces}, {functions}, {classes}, and
{namespaces} are stored separately. Function arguments are
correctly described, identifying complex types like union types or
callbacks. Optional parameters are also identified. For
{classes}, a distinction between constructors and methods
is made. Declaration files are tagged according to their features,
which is useful for identifying and filtering out declaration files
that contain unimplemented features.

\texttt{dts-parse} is itself written in TypeScript. It receives the file as input
parameter and returns an AST serialized in JSON.

% , as shown in \coderef{code:dts-parse-example}.
% \begin{lstlisting}[
%   caption={Example for \texttt{dts-parse} executed on module \texttt{glob-to-regexp} and combined with \texttt{jq} to browse the returned JSON},
%   language=bash,
% label=code:dts-parse-example,
%   float=tp, captionpos=b
% ]
% $ dts-parse -i glob-to-regexp/index.d.ts | jq '.parsing.functions[0].name'
% "GlobToRegexp"
% \end{lstlisting} %$

\subsection{dts-compare}
\label{sec:dts-compare}

% The tool is written in TypeScript.
The compare tool takes two declaration files as input and returns a
JSON file containing the result of the comparison as shown in
\coderef{code:dts-compare-example}. Following the standard naming convention of unit
testing frameworks, the input files are marked as \lstinline{expected} or
\lstinline{actual}. We use \lstinline{expected} for the DefinitelyTyped
declaration files.

@@

\texttt{dts-compare} applies \texttt{dts-parse} internally to both files to get 
normalized structures that are easy to compare. The result of the comparison is an array
of an abstract entity \texttt{Difference}. 
A difference is \texttt{solvable} if it can be resolved by adding more code
examples. For example, invoking a function with different parameters so that a missing
interface property gets accessed. Otherwise it is  \texttt{unsolvable}.

The concrete evaluated differences are as follows:
\begin{itemize}
  \item \texttt{TemplateDifference}: A difference between the choice of TypeScript
    templates. For example, if the file in DefinitelyTyped is written as a \textbf{module}
    but we generated the file using the \textbf{module-function} template. The comparison
    stops if the templates differ. 
  \item \texttt{ExportAssignmentDifference}: An equality check between the export
    assignments of both files, that is in the expression
    \lstinline{export = XXX}.
    \begin{changethis}
      Q: how severe is that? That's not an issue, right?
    \end{changethis}
  \item \texttt{FunctionMissingDifference}: A function declaration is present in the expected
    file but not in the generated one. This difference applies to functions as well as methods of classes
    and properties of interfaces. 
  \item \texttt{FunctionExtraDifference}: A function is present in the generated declaration file but not in DefinitelyTyped.
  \item \texttt{FunctionOverloadingDifference}: The number of declarations for the same
    function is different. This difference can be due to inexperience of the author of a
    declaration file, who uses, e.g., multiple declarations where a union typed argument
    would be appropriate.
  \item \texttt{ParameterMissingDifference}: A parameter of a function or a property of an interface is not present in the generated file.
  \item \texttt{ParameterExtraDifference}: A parameter of a function or a property of an
    interface is present in the generated file but not in the DefinitelyTyped file. 
  \item \texttt{ParameterTypeDifference}: A parameter of a function or a property of an interface is generated with a different type than in the DefinitelyTyped file. Here we differentiate between \texttt{SolvableDifference} and \texttt{UnsolvableDifference}.
  \begin{itemize}
    \item \texttt{SolvableDifference}: A type difference that can be solved by writing
      further code examples. For example, a basic type that is converted to a union type,
      a function overloading, or a parameter or property that is marked as optional. 
    \item \texttt{UnsolvableDifference}: Any difference not considered as \texttt{SolvableDifference}.
  \end{itemize}
\end{itemize}

@@

\begin{changethis}

To measure the impact of incomplete code examples on the
quality of generated declaration files we categorized each difference as follows:

  We extracted code examples written by developers to execute the code
  and gather run-time information. The quality of the generated
  declaration file is heavily dependent on the code examples. For
  example, if a property of an interface does not get accessed due to
  a code example not exploring a particular execution path, that
  property will not exist in the generated declaration file.
\end{changethis}

@@

\paragraph*{Extraction of Code Examples}
In this step, we loose another 50\% of modules! This loss is mainly
explained because some developers do not wrap their code in a block
using the \texttt{javascript} or \texttt{js} tags. However, as we are
still left with code examples for 2260 modules, we did not further
look into code extraction as this number was considered sufficient for
evaluating the generation of declaration files.
